New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-09-11 jso.qmd
finished revising the Practicalities section

title: Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

@@

The paper is structured as follows. @sec-background introduces the
background of the PPGT, reviews existing optimisers and index
functions in the literature. @sec-theory describes JSO and introduces the 
metrics that measure different properties of PP indexes, smoothness and
squintability. @sec-sim-deets summarises the results of two simulation experiments done to assess JSO's performance: one comparing JSO's performance improvements relative to an existing optimiser, Creeping Random Search (CRS), and the other studying the impact of different PP index properties on optimisation performance. @sec-sim-res presents the results. @sec-discussion discusses implementation details and insights for practitioners. @sec-conclusion summarises the work and provides suggestions for future
directions.

@@

A tour on high-dimensional data is constructed by geodesically
interpolating between pairs of planes. Any plane is described by an
orthonormal basis, $A_t$, where $t$ represents time in the sequence. The
term "geodesic" refers to maintaining the orthonormality constraint so
that each view shown is correctly a projection of the data. The PP
guided tour operates by geodesically interpolating to target planes
(projections) which have high PP index values, as provided by the
optimiser. The geodesic interpolation means that the viewer sees a
continuous sequence of projections of the data, so they can watch
patterns of interest forming as the function is optimised. There are
five (unsophisticated) optimisation methods implemented in the `tourr`
package:

-   `search_geodesic()`: provides a pseudo-derivative optimisation. It
    searches locally for the best direction, based on differencing the
    index values for very close projections. Then it follows the
    direction along the geodesic path between planes, stopping when the
    next index value fails to increase.
-   `search_better()`: also known as Creeping Random Search (CRS), is a
    brute-force optimisation searching randomly for projections with
    higher index values.
-   `search_better_random()`: is essentially simulated annealing
    [@Bertsimas93] where the search space is reduced as the optimisation
    progresses.
-   `search_posse()`: implements the algorithm described in @posse95.
-   `search_polish()`: is a very localised search, to take tiny steps to
    get closer to the local maximum.

There are several PP index functions available: `holes()` and `cmass()`
[@cook1993projection]; `lda_pp()` [@lee2005projection]; `pda_pp()`
[@lee2010projection]; `dcor2d()` and `splines2d()` [@Grimm2016];
`norm_bin()` and `norm_kol()` [@huber85]; `slice_index()`
[@Laa:2020wkm]. Most are relatively simply defined, for any projection
dimension, and implemented because they are relatively easy to optimise.
A goal is to be able to incorporate more complex PP indexes, for
example, based on scagnostics (@scag, @WW08).

@@

# The jellyfish optimiser and properties of PP indexes {#sec-theory}

@@

@@

The index properties, including smoothness and squintability, offer
numerical metrics to characterise the complexity of projection pursuit
optimisation problems. @tbl-smoothness-squintability presents the
parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function ($\theta_1$ to $\theta_4$) for each case considered in the
simulation. The column "smooth" is used as the smoothness metric and the
column "squint" is calculated as equation (@eq-squintability-parametric)
as the squintability metric. @tbl-mod-output presents the results of
fitting a generalised linear model with a binomial family and a logit
link function to a sample of data in @tbl-mod-data, where all the three
components (success rate, JS hyper-parameters, and index properties) are
combined. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the index
properties: smoothness and squintability. Specifically, using 10 more
jellyfish and 10 more tries increases the odd ratio of success by
`r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`% and
`r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success.

@@

```{r}
#| label: tbl-mod-data
#| tbl-cap: The first 7 rows of the datasets processed for modelling. 
sim_df |> 
  dplyr::select(index, d, smoothness, 
    squintability, n_jellies, max_tries, 
    -I_max, P_J, time) |> 
  mutate(time = as.numeric(time)) |>
  arrange(index, d, n_jellies, max_tries) |> 
  head(7) |> 
  kable(digits = 2, format = "latex", 
        col.names = c("index", "d", 
          "smoothness", "squintability", 
          "n. jellyfish", 
          "max. tries", 
          "success rate", "time (sec)"),
        linesep = "",
        booktabs = T) |> 
  kable_styling(#latex_options = "hold_position",
    font_size = 10)
```

@@

The simulation study illustrates how these metrics can be used to assess a PP index, and the optimisation process. In general, we would expect that the JSO is going to outperform most other optimisers. From an exploratory perspective, it take a shift in the GTPP practice because one no longer has a single path to follow. Here we describe how to adjust to using the JSO optimiser for exploring high-dimensional data, and how to use the new metrics for evaluating a new index.

@@

To use JSO for PPGT in the `tourr` package, specify `search_f = search_jellyfish` in the guided tour. Since multiple tour paths are explored by the jellyfish, the animation function `animate_*()` will not render the tour path for JSO. To visualise path visited by each individual jellyfish, assign the animation to an object (`res <- animate_xy(...)`). This will save the bases visited by the optimiser, along with relevant metadata, as a tibble data object (see @RJ-2021-105 for more details on the data object). The bases visited for individual jellyfish can then be extracted and viewed as a planned tour. The following code uses JSO, with 20 jellyfish and a maximum of 50 tries, on the penguins data using the `lda_pp` index.

```{r}
#| echo: true
#| eval: false
# clean up the penguins data
penguins_cl <- palmerpenguins::penguins |> 
  filter(!is.na(bill_length_mm)) |>
  rename(bl = bill_length_mm, bd = bill_depth_mm,
         fl = flipper_length_mm, bm = body_mass_g)
penguins_df <- penguins_cl |> select(bl:bm) |> rescale()
# run the projection pursuit with JSO 
res <- animate_xy(penguins_df, 
                  guided_tour(lda_pp(cl = penguins_cl$species),
                              search_f = search_jellyfish, 
                              n_jellies = 20, max.tries = 50))
```

@@

```{r}
#| echo: true
#| eval: false
# extract the bases from the first jellyfish
bases <- res |> filter(loop == 1) |> pull(basis) |> check_dup(0.1)
# visualise its path with a planned tour
animate_xy(data = penguins_df, tour_path = planned_tour(bases), 
           col = penguins_cl$species)
```

@fig-jso-penguins shows six frames from the tour jellyfish No.1 animation. The Gentoo species (red) was mixed with Chinstrap (yellow) initially and gradually separated out as flipper length (`fl`) gaining more weights.

@@

```{r}
#| label: fig-jso-penguins
#| fig-cap: "Visualising the tour path of a single jellyfish (No.1) in the penguins data set. The projection pursuit is optimised by the JSO with 20 jellyfishes and a maximum of 50 tries. The bases visited by jellyfish No.1 is extracted and constructed as a planned tour. The jellyfish successfully finds the structure of the data set, separating the three species of penguins."
frames <- c("001", "010", "019", "028", "036", "048")
ani <- paste0(here::here("figures/"), "tour/", "penguins", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl)
```

@@

The `ferrn` package provides functionality for computing the smoothness and squintability metrics. Both cases require sampling random bases using the `sample_bases()` function, followed by  calculating the specific metric with either `calc_smoothness()` or `calc_squintability()`.

@@

To sample bases for calculating smoothness, the required arguments include `idx` (the index function used), `data` (the dataset), `n_basis` (the number of random bases to sample), and `best` (the best basis to compute projection distance). The following code samples 300 bases for using the `holes` index to find the pipe shape in the 6D `sine1000` dataset:

```{r}
#| echo: true
library(GpGp)
library(fields)
library(tourr)
library(ferrn)
basis_smoothness <- sample_bases(
  idx = "holes", data = sine1000, n_basis = 300, 
  best = matrix(c(0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 1), nrow = 6))
basis_smoothness |> head(3)
```

@@

The `calc_smoothness()` function takes the output from `sample_bases()` and fits a Gaussian process model to the index values against the projection distances to compute the smoothness metric. The starting parameter for the fit can be specified using the `start_params` argument, and additional Gaussian process arguments can be provided through `other_gp_params`. The following code calculates the smoothness metric from the previously sampled bases:

```{r}
#| echo: true
(sm_res <- calc_smoothness(basis_smoothness))
```

@@

The `calc_squintability()` function computes squintability using two methods: 1) parametrically, by fitting a scaled sigmoid function through non-linear least square (`method = "nls"`), and 2) non-parametrically, using kernel smoothing (`method = "ks"`). A `bin_width` argument is required to average the index values over the projection distance before the fit. The following code samples 100 bases for calculating squintability in the same scenario:

```{r}
#| echo: true
basis_squint <- sample_bases(
   idx = "holes", data = sine1000, n_basis = 100, 
  best = matrix(c(0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 1), nrow = 6),
  step_size = 0.1, min_proj_dist = 1.5)
calc_squintability(basis_squint, method = "nls", bin_width = 0.01)
```

@@

The article is created using Quarto [@Allaire_Quarto_2022] in R [@R].
The source code for reproducing the work reported in this paper can be
found at: <https://github.com/huizezhang-sherry/paper-jso>. Nicolas
Langrené acknowledges the partial support of the Guangdong Provincial
Key Laboratory IRADS (2022B1212010006, R0400001-22) and the UIC Start-up
Research Fund UICR0700041-22.

>>>>

title: New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

@@

The paper is structured as follows. @sec-background introduces the
background of the PPGT, reviews existing optimisers and index
functions in the literature. @sec-PP-properties introduces the 
metrics that measure different properties of PP indexes, smoothness and
squintability, and @sec-JSO describes the new JSO to be used for the PPGT. @sec-sim-deets summarises the results of two simulation experiments done to assess JSO's performance: one comparing JSO's performance improvements relative to an existing optimiser, Creeping Random Search (CRS), and the other studying the impact of different PP index properties on optimisation performance. @sec-sim-res presents the results. @sec-discussion discusses implementation details and insights for practitioners. @sec-conclusion summarises the work and provides suggestions for future
directions.

@@

A tour on high-dimensional data is constructed by geodesically
interpolating between pairs of planes. Any plane is described by an
orthonormal basis, $A_t$, where $t$ represents time in the sequence. The
term "geodesic" refers to maintaining the orthonormality constraint so
that each view shown is correctly a projection of the data. The PP
guided tour operates by geodesically interpolating to target planes
(projections) which have high PP index values, as provided by the
optimiser. The geodesic interpolation means that the viewer sees a
continuous sequence of projections of the data, so they can watch
patterns of interest forming as the function is optimised. There are
five optimisation methods implemented in the `tourr`
package:

- a pseudo-derivative, that searches locally for the best direction, based on differencing the index values for very close projections.
- a brute-force optimisation (CRS). 
- a modified brute force algorithm described in @posse95. 
- an essentially simulated annealing [@Bertsimas93] where the search space is reduced as the optimisation.
- a very localised search, to take tiny steps to get closer to the local maximum.

<!-- 
-   `search_geodesic()`: provides a pseudo-derivative optimisation. It
    searches locally for the best direction, based on differencing the
    index values for very close projections. Then it follows the
    direction along the geodesic path between planes, stopping when the
    next index value fails to increase.
-   `search_better()`: also known as Creeping Random Search (CRS), is a
    brute-force optimisation searching randomly for projections with
    higher index values.
-   `search_better_random()`: is essentially simulated annealing
    [@Bertsimas93] where the search space is reduced as the optimisation
    progresses.
-   `search_posse()`: implements the algorithm described in @posse95.
-   `search_polish()`: is a very localised search, to take tiny steps to
    get closer to the local maximum.
-->

There are numerous PP index functions available: introduced in @cook1993projection, @lee2005projection, @lee2010projection, @Grimm2016, @huber85, @Laa:2020wkm. Most are relatively simply defined, for any projection dimension, and implemented because they are relatively easy to optimise. A goal is to develop PP indexes based on scagnostics (@scag, @WW08), but the blockage is optimisation that these tend to be noisy, with potentially small squint angles.

@@

# Properties of PP indexes {#sec-PP-properties}

@@

# The jellyfish optimiser {#sec-JSO}

@@

The index properties, including smoothness and squintability, offer
numerical metrics to characterise the complexity of projection pursuit
optimisation problems. @tbl-smoothness-squintability presents the
parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function ($\theta_1$ to $\theta_4$) for each case considered in the
simulation. The column "smooth" is used as the smoothness metric and the
column "squint" is calculated as equation (@eq-squintability-parametric)
as the squintability metric. @tbl-mod-output presents the results of
fitting a generalised linear model with a binomial family and a logit
link function.<!--, where all the three
components (success rate, JS hyper-parameters, and index properties) are
combined.--> The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the index
properties: smoothness and squintability. Specifically, using 10 more
jellyfish and 10 more tries increases the odd ratio of success by
`r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`% and
`r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success.

@@

<!-- 
```{r}
#| label: tbl-mod-data
#| tbl-cap: Data constructed to store simulation results (the first 7 rows) includes name of index, dimension, index metrics, JSO parameters, and performance metrics. 
sim_df |> 
  dplyr::select(index, d, smoothness, 
    squintability, n_jellies, max_tries, 
    -I_max, P_J, time) |> 
  mutate(time = as.numeric(time)) |>
  arrange(index, d, n_jellies, max_tries) |> 
  head(7) |> 
  kable(digits = 2, format = "latex", 
        col.names = c("index", "d", 
          "smoothness", "squintability", 
          "n. jellyfish", 
          "max. tries", 
          "success rate", "time (sec)"),
        linesep = "",
        booktabs = T) |> 
  kable_styling(#latex_options = "hold_position",
    font_size = 10)
```
-->

@@

The simulation study demonstrated the that these metrics can be used to assess a PP index, and compare optimisers. Here we explain how they can be used for new index and optimiser development. Using the JSO optimiser for GTPP requires a change in user behaviour. For all the existing optimisers a single optimisation path is followed. The JSO has multiple optimisation paths, and needs additional tools to incorporate into the PPGT, as explained here.

@@

To use JSO for PPGT in the `tourr` package, specify `search_f = search_jellyfish` in the guided tour. Since multiple tour paths are generated by the jellyfish, the animation function `animate_*()` will not render the tour path for JSO. To visualise path visited by each individual jellyfish, assign the animation to an object (`res <- animate_xy(...)`). This will save the bases visited by the optimiser, along with relevant metadata, as a tibble data object (see @RJ-2021-105 for more details on the data object). The bases visited for individual jellyfish can then be extracted and viewed using the `planned_tour()` function. <!-- The following code uses JSO, with 20 jellyfish and a maximum of 50 tries, on the penguins data using the `lda_pp` index.-->

```{r}
#| echo: false
#| eval: false
# clean up the penguins data
penguins_cl <- palmerpenguins::penguins |> 
  filter(!is.na(bill_length_mm)) |>
  rename(bl = bill_length_mm, bd = bill_depth_mm,
         fl = flipper_length_mm, bm = body_mass_g)
penguins_df <- penguins_cl |> select(bl:bm) |> rescale()
# run the projection pursuit with JSO 
res <- animate_xy(penguins_df, 
                  guided_tour(lda_pp(cl = penguins_cl$species),
                              search_f = search_jellyfish, 
                              n_jellies = 20, max.tries = 50))
```

@@

```{r}
#| echo: false
#| eval: false
# extract the bases from the first jellyfish
bases <- res |> filter(loop == 1) |> pull(basis) |> check_dup(0.1)
# visualise its path with a planned tour
animate_xy(data = penguins_df, tour_path = planned_tour(bases), 
           col = penguins_cl$species)
```

<!-- @fig-jso-penguins shows six frames from the tour jellyfish No.1 animation. The Gentoo species (red) was mixed with Chinstrap (yellow) initially and gradually separated out as flipper length (`fl`) gaining more weights.-->

@@

<!-- 
```{r}
#| label: fig-jso-penguins
#| fig-cap: "Visualising the tour path of a single jellyfish (No.1) in the penguins data set. The projection pursuit is optimised by the JSO with 20 jellyfishes and a maximum of 50 tries. The bases visited by jellyfish No.1 is extracted and constructed as a planned tour. The jellyfish successfully finds the structure of the data set, separating the three species of penguins."
frames <- c("001", "010", "019", "028", "036", "048")
ani <- paste0(here::here("figures/"), "tour/", "penguins", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl)
```
-->

@@

The `ferrn` package [@RJ-2021-105] provides functionality for computing the smoothness and squintability metrics. Both cases require sampling random bases using the `sample_bases()` function, followed by  calculating the specific metric with either `calc_smoothness()` or `calc_squintability()`.

@@

To sample bases for calculating smoothness, the required arguments include `idx` (the index function used), `data` (the dataset), `n_basis` (the number of random bases to sample), and `best` (the best basis to compute projection distance). <!--The following code samples 300 bases for using the `holes` index to find the pipe shape in the 6D `sine1000` dataset: -->

```{r}
#| echo: false
#| eval: false
library(GpGp)
library(fields)
library(tourr)
library(ferrn)
basis_smoothness <- sample_bases(
  idx = "holes", data = sine1000, n_basis = 300, 
  best = matrix(c(0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 1), nrow = 6))
basis_smoothness |> head(3)
```

@@

The `calc_smoothness()` function takes the output from `sample_bases()` and fits a Gaussian process model to the index values against the projection distances to compute the smoothness metric. The starting parameter for the fit can be specified using the `start_params` argument, and additional Gaussian process arguments can be provided through `other_gp_params`. <!--The following code calculates the smoothness metric from the previously sampled bases:-->

```{r}
#| echo: false
#| eval: false
(sm_res <- calc_smoothness(basis_smoothness))
```

@@

The `calc_squintability()` function computes squintability using two methods: 1) parametrically, by fitting a scaled sigmoid function through non-linear least square (`method = "nls"`), and 2) non-parametrically, using kernel smoothing (`method = "ks"`). A `bin_width` argument is required to average the index values over the projection distance before the fit. <!-- The following code samples 100 bases for calculating squintability in the same scenario:-->

```{r}
#| echo: false
#| eval: false
basis_squint <- sample_bases(
   idx = "holes", data = sine1000, n_basis = 100, 
  best = matrix(c(0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 1), nrow = 6),
  step_size = 0.1, min_proj_dist = 1.5)
calc_squintability(basis_squint, method = "nls", bin_width = 0.01)
```

@@

The article is created using Quarto [@Allaire_Quarto_2022] in R [@R].
The source code for reproducing the work reported in this paper can be
found at: [https://github.com/huizezhang-sherry/paper-jso](https://github.com/huizezhang-sherry/paper-jso). Nicolas
Langrené acknowledges the partial support of the Guangdong Provincial
Key Laboratory IRADS (2022B1212010006, R0400001-22) and the UIC Start-up
Research Fund UICR0700041-22.
